Tidy data structure to support exploration and modeling of temporal-context data

### 2018-08-12 main.Rmd
rewrite the intro to ts

Temporal-context data consist of observational units indexed at different time points $X_{jt}$, where the $j$\textsuperscript{th} unit takes measurements of $X$ on over time $t$, for $j = 1, \dots, N_{i}$ and $1 \le t \le T$. Time primarily forms the contextual basis of temporal data, but it could arrive in many possible formats. For example, data recorded at fine time resolutions (hours, minutes, and seconds) are typically associated with different time zones and daylight savings. Temporal data often carries with rich information other than the time: multiple observational units of different time lengths, multiple and heterogeneous measured variables, multiple grouping factors involving nested or crossed structures, linking to other data tables, and etc.

In the literature, time series and panel (longitudinal) data are common terms referred to temporal-context data, depending on the research fields. Researchers who are concerned with modelling large $T$ and small $N$ would name as "time series" (serial correlation); those who are interested in modelling small $T$ and large $N$ as "panel data" (asymptotic). The data format is two-dimensional array, but different modelling focuses lend the data input to different representations. A matrix is used to represent multivariate time series where each row represents observations measured at a time point and each column represents a series ("wide" form). This matrix representation requires homogeneity (that is, all the columns must be of same type.), time indices implicitly inferred as attributes or meta-information, series of same time length, and explicit missingness. By contrast, panel data are organised in rectangular form of heterogeneous column types where multiple study subjects are stacked and repeated for its time indices in a single column ("long" form), due to commonly arisen unbalanced panels. This specification requires explicitly declared panel variable and index, which has been implemented in Stata's time series module and R package **plm** [@R-plm]. Evidently, this data organisation appears more flexible than matrix in supporting explicit time index, multiple subjects of different time lengths, and implicit missing values. less restrictions to be made. [R]

Temporal data can often be aggregated in a manner that exhibits a nested or crossed structure, also known as hierarchical or grouped time series [@fpp]. For example, in a manufacturing setting, a company can add up every store’s sales by region, by state and by country, which gives a strictly hierarchical time series; alternatively they can group the sales for each product together based on common attributes such as store, brand, price range and so forth, which leads to a non-hierarchical structure---a grouped time series. The R package **hts** [@R-hts] is the implementation of this type of time series. But it is frustrating to create and work with such data objects due to non-extensible and non-modified. Variational data aggregation (some groups no need aggregation). [R]

@wickham2014tidy coined the term "tidy data" and formalized the processing from messy data to "tidy data". These principles attempt to standardize the mapping from the semantics of a dataset to its structure and facilitate data analysis in a coherent way. Based on the systematic structuring principles, the grammar of data manipulation, as implemented in the R package **dplyr** [@R-dplyr], abstracts an coherent and consistent set of methods to handle a wide range of data transformation tasks. @r4ds argues that 80% of data analysis tasks can be solved with tidy tools while the remaining 20% requires other tools. [R]

@@

The choice of representation of temporal-context data is made from a data-centric perspective, which is taken in the light of the operations that are to be performed on the data. This data abstraction reflects most common problems of incoming data and transformation in reality. Firstly, a data set must be structured in a "tidy" rectangular layout each row has the same schema and each field has a single value. Secondly, declaring the data set to contain temporal observations is determined by an "index" that represents time and a "key" that uniquely identifies each unit that measurements take place on over time. The "key" serves a similar purpose as the panel variable in the Stata's **tsset** command to define the units or subjects, but it is expanded to include multiple variables rather than a single one. A syntax is introduced to express a key consisting of nested and crossed factors. The composition of index and key uniquely defines each observation in a data table, which is equivalent to a primary key [@codd_relational_1970] in a relational database.

@@

The "key" identifies units or subjects that are observed over time in a data table, which are typically known in advance by users. The absence of key only occurs to a univariate case where the key is implicit. Multiple units must associate with an explicit "key" to allow for identification. The key is not constrained to a single column, but is comprised of multiple columns. When involving multiple variables, one need to distinguish nesting and crossing data variables. "Nesting" refers to a scenario where a variable is nested within another when each category of the former variable co-occurs with only one category of the latter (for example, `country` nested within `continent`); on the other hand, "crossing" means that a variable is crossed with another when every category of one variable co-occurs with every category of the other (for example, `country` crosses with `gender`). Nesting is a special case of crossing. The former exhibits a hierarchical ordering with lower-level category (children) nested within higher-level category (parent) (more categories in lower level); whereas the ordering of the latter makes no difference. Whilst constructing tidy temporal data, users also need to specify the key that identifies the structure on the data. Rather than creating a separate object that contains the data structure, like what the **hts** does, we propose a syntax-based interface to directly deal with variables, so that it takes advantage of data semantics. The expression of a vertical bar (`|`) (conditional) is used to indicate nesting variables, while a comma (`,`) for crossing variables. @wilkinson2006grammar discussed the distinction between nesting and crossing with respect to facets in graphics, and he used `/` and `*` instead for nested and crossed expressions. A crossing structure displays all unique and possible combinations of crossed variables, including those not found in the data, in a graphical layout. However, tidy temporal data finds only the combinations that occur in the data and intentionally respects the structure of incomplete combinations.

@@

What is notable about data pipeline is that it not only frees users from a tedious and error-prone analysis but also empowers them to focus on the statistical analysis tasks at hand without concerning the details of computational implementation. A fluent and coherent pipeline glues the grammar of data manipulation and tidy data as a fundamental unit of data analysis together. It helps (1) break up a big problem to into manageable blocks, (2) generate human readable analysis workflow, (3) avoid introducing mistakes as many as possible.

@@

A new visualization framework for handling time series with nesting and crossing structure is under development. The proposed data structure proves a natural input that can be piped into a visualization program. As long as the program understands the embedded structure, it will produce sensible plots.

@@

```{r write-bib}
write_bib(c("hts", "plm", "dplyr"), file = "rpkgs.bib")
```

>>>>

Temporal-context data consist of observational units indexed at different time points $X_{jt}$, where the $j$\textsuperscript{th} in advance unit takes measurements of $X$ on over time $t$, for $j = 1, \dots, N_{i}$ and $1 \le t \le T$. Time primarily forms the contextual basis of temporal data, but it could arrive in many possible formats. For example, data recorded at fine time resolutions (hours, minutes, and seconds) are typically associated with different time zones and daylight savings. Temporal data often carries with rich information other than the time: multiple observational units of different time lengths, multiple and heterogeneous measured variables, multiple grouping factors involving nested or crossed structures, linking to other data tables, and etc.

## Time series and longitudinal data

Time series and longitudinal data are the common terms when referring to temporal-context data. Despite of exactly the same data input, its representation diverges in these two areas due to different modelling focuses. Time series can be univariate or multivariate, but usually involves observations of relatively large in length possessed by each series. Time series researchers and analysts who are concerned with this large $T$ property, are mostly interested in modelling temporal dependency. Time series are represented as vectors or matrices in statistical software, for example MATLAB [@Matlab] and R [@R-base] (including `ts`, **zoo** [@zoo2005], and **xts** [@R-xts]). In the matrix case, each row is assumed to contain observations measured at a time point and each column to contain a single time series. This implies that data columns of homogeneous types: numeric or non-numeric, but there are limited supporting methods for non-numeric cases. In addition to homogeneity, time indices must be stripped off the data and implicitly inferred as attributes or meta-information; the number of observations must be the same across all the series. Data wrangling from "raw" form to this specialist format can be frustrating and difficult, in order to perform a variety of downstream tasks such as analytics.

On the other hand, those who are interested in modelling small $T$ and large $N$ as "panel data" (asymptotic). By contrast, panel data are organised in rectangular form of heterogeneous column types where multiple study subjects are stacked and repeated for its time indices in a single column ("long" form), due to commonly arisen unbalanced panels. This specification requires explicitly declared panel variable and index, which has been implemented in Stata's time series module and R package **plm** [@R-plm]. Evidently, this data organisation appears more flexible than matrix in supporting explicit time index, multiple subjects of different time lengths, and implicit missing values. less restrictions to be made. [R]

## Hierarchical and grouped time series

Temporal data can often be aggregated in a manner that exhibits a nested or crossed structure, also known as hierarchical or grouped time series [@fpp]. For example, in a manufacturing setting, a company can add up every store’s sales by region, by state and by country, which gives a strictly hierarchical time series; alternatively they can group the sales for each product together based on common attributes such as store, brand, price range and so forth, which leads to a non-hierarchical structure---a grouped time series. Additionally, hierarchical structure can be derived from statistical models that classifies the time series. The R package **hts** [@R-hts] is the implementation of this type of time series. But it is frustrating to create and work with such data objects due to non-extensible and non-modified. Variational data aggregation (some groups no need aggregation). [R]

## Tidy data

@wickham2014tidy coined the term "tidy data", which is a rephrasing of the second and third normal forms in relational databases. and formalized the processing from messy data to "tidy data". These principles attempt to standardize the mapping from the semantics of a dataset to its structure and facilitate data analysis in a coherent way. Based on the systematic structuring principles, the grammar of data manipulation, as implemented in the R package **dplyr** [@R-dplyr], abstracts an coherent and consistent set of methods to handle a wide range of data transformation tasks. @r4ds argues that 80% of data analysis tasks can be solved with tidy tools while the remaining 20% requires other tools. [R]

@@

The choice of representation of temporal-context data is made from a data-centric perspective, which is taken in the light of the operations that are to be performed on the data. This data abstraction reflects most common problems of incoming data and transformation in reality. Firstly, a data set must be structured in a "tidy" rectangular layout each row has the same schema and each field has a single value. Secondly, declaring the data set to contain temporal observations is determined by an "index" that represents time and a "key" that uniquely identifies each unit that measurements take place on over time. The "key" serves a similar purpose as the panel variable in the Stata's **tsset** command to define the units or subjects, but it is expanded to include multiple variables rather than a single one. A syntax is introduced to express a key consisting of nested and crossed factors. The composition of index and key uniquely defines each observation in a data table, which is equivalent to a primary key [@codd_relational_1970] in a relational database. (The matrix representation also implies that every observation is determined by column variable and time-indexed row.)

@@

The "key" identifies units or subjects that are observed over time in a data table, which is typically a priori known structure by users. The absence of key only occurs to a univariate case where the key is implicit. Multiple units must associate with an explicit "key" to allow for identification. The key is not constrained to a single column, but is comprised of multiple columns. When involving multiple variables, one need to distinguish nesting and crossing data variables. "Nesting" refers to a scenario where a variable is nested within another when each category of the former variable co-occurs with only one category of the latter (for example, `country` nested within `continent`); on the other hand, "crossing" means that a variable is crossed with another when every category of one variable co-occurs with every category of the other (for example, `country` crosses with `gender`). Nesting is a special case of crossing. The former exhibits a hierarchical ordering with lower-level category (children) nested within higher-level category (parent) (more categories in lower level); whereas the ordering of the latter makes no difference. Whilst constructing tidy temporal data, users also need to specify the key that identifies the structure on the data. Rather than creating a separate object that contains the data structure, like what the **hts** does, we propose a syntax-based interface to directly deal with variables, so that it takes advantage of data semantics. The expression of a vertical bar (`|`) (can be interpreted probabilistically as the conditional distribution for its parent) is used to indicate nesting variables, while a comma (`,`) for crossing variables. @wilkinson2006grammar discussed the distinction between nesting and crossing with respect to facets in graphics, and he used `/` and `*` instead for nested and crossed expressions. A crossing structure displays all unique and possible combinations of crossed variables, including those not found in the data, in a graphical layout. However, tidy temporal data finds only the combinations that occur in the data and intentionally respects the structure of incomplete combinations.

@@

What is notable about data pipeline is that it not only frees users from a tedious and error-prone analysis but also empowers a wider audience to focus on the statistical analysis tasks at hand without concerning the details of computational implementation. A fluent and coherent pipeline glues the grammar of data manipulation and tidy data as a fundamental unit of data analysis together. It helps (1) break up a big problem to into manageable blocks, (2) generate human readable analysis workflow, (3) avoid introducing mistakes as many as possible.

@@

A new visualization framework for handling time series with nesting and crossing structure is under development. The proposed data structure proves a natural input (unit time series and parent-children structure) that can be piped into a visualization program. As long as the program understands the embedded structure, it will produce sensible layout and interactivity.

@@

```{r write-bib}
write_bib(c("hts", "plm", "dplyr", "xts", "zoo"), file = "rpkgs.bib")
```
